---
title: "Class10"
author: "Gregory Jordan"
format: pdf
editor: visual
toc: true
---


# Background

In this mini project we wille xamine 538 Halloween candy data. 
what is your favorite candy? what is nougat anyway? and how do you say it in america?

# 1. Importing the dataset

First step is to read the data:

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(candy_file,row.names = 1)
head(candy)
```
win percent means this candy was their favorite when asked people

Q1. How many different candy types are in this dataset?
```{r}
cat(nrow(candy),"different types of candy")
```

Q2. How many fruity candy types are in the dataset?

```{r}
cat(sum(candy$fruity),"fruity candy types in the dataset")
```
# 2. What is your favorite candy?

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
#let's see all the different types of candy to find my favorite. 
row.names(candy)
```
```{r}
#my favorite candy is Twix
candy["Twix",]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
#install.packages("skimr")
library(skimr)
skimr::skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent is on a 1-100% scale while the other variables are on a 0-1 scale.

Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero and one represent whether the candey was a chocolate type of candy or not

Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```
```{r}
#we can use ggplot also to plot the hist and make it fancier
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(bins=10,col="red",fill="yellow")
```

Q9. Is the distribution of winpercent values symmetrical?

No. skewed towards <50%

Q10. Is the center of the distribution above or below 50%?

Center of the distribution is skewed below 50%

Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.win<- mean(candy$winpercent[as.logical(candy$chocolate)])
cat(chocolate.win, "= mean win % chocolate\n")

fruity.win<- mean(candy$winpercent[as.logical(candy$fruity)])
cat(fruity.win,"= mean win $ fruity\n")

cat("Chocolate > Fruity?", chocolate.win>fruity.win)

```

Q12. Is this difference statistically significant?

```{r}
#student t test to test for significance
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```
yes it is stat significant b/c super low p value

# 3. Overall Candy Rankings

the base R `sort()` and `order()` functions are very useful! dplyr works well too!

Q13. What are the five least liked candy types in this set?

```{r}
#I like tidyverse/dplyr
library(dplyr)
candy.least.liked <- candy %>% arrange(winpercent)
head(candy.least.liked,5)
```

```{r}
#can also use base R instead of tidyverse
inds <- order(candy$winpercent)
head(candy[inds,],5)
```

Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
#using base R
inds <- order(candy$winpercent)
tail(candy[inds,],5)
```

Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
#library(ggplot2)
ggplot(data=candy) + aes(winpercent,rownames(candy)) + geom_col()

```
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(data=candy) + aes(winpercent,reorder(rownames(candy),winpercent)) + geom_col()
```
you can use `ggsave()` to save/edit dimensions and save your most recent plot if you want

let's add color. use a color vector that we can then use to color candies by descriptions (like chocolate and stuff)

```{r}
#start by making a vector of all black color as long as the different types of candies
my_cols <- rep("black",nrow(candy))
#my_cols
#then overwrite the vector to rename colors based off candy
my_cols[as.logical(candy$chocolate)] <- "darkgreen"
my_cols[as.logical(candy$bar)] <- "blue"
my_cols[as.logical(candy$fruity)] <- "red"
my_cols
```

```{r}
ggplot(data=candy) + aes(winpercent,reorder(rownames(candy),winpercent)) + geom_col(fill=my_cols)
```
Q17. What is the worst ranked chocolate candy?

sixlets

```{r}
candy.worst.chocolate <- candy %>% filter(chocolate==1) %>% arrange(winpercent)
candy.worst.chocolate[1,]
```


Q18. What is the best ranked fruity candy?

starburst 

```{r}
candy.best.fruity <- candy %>% filter(fruity==1) %>% arrange(desc(winpercent))
candy.best.fruity[1,]
```


# 4. Taking a look at pricepercent

what about value for money?
what is the besat candy for the least money?

one way to get at this would be to make a plot of `winpercent` vs the `pricepercent` variables

```{r}
#install.packages(ggrepel)
#ggrepel to make labels not overlap
library(ggrepel)
ggplot(candy) + aes(winpercent,pricepercent,label=rownames(candy)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, size=3.3, max.overlaps = 7)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
candy.bangforbuck <- candy %>% mutate(bangforbuck=winpercent/pricepercent) %>% arrange(desc(bangforbuck))
candy.bangforbuck
```
tootsie roll midgies

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy.pricesorted <- candy %>% arrange(desc(pricepercent))
candy.pricesorted[1:5,] %>% arrange(winpercent)
```
Nik L Nip is least popular of top 5 most expensive candies

# 5. Exploring the correlation structure

```{r}
#install.packages("corrplot")
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot::corrplot(cij)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity (not really many fruity chocolate candies at all)

Q23. Similarly, what two variables are most positively correlated?

chocolate and bar and chocolate and winpercent (lots of chocolate bars and people like chocolate)

# 6. PCA: Principal Component Analysis

the main function that always there for us is `prcomp()`. it has an important argument that is set to `scale=FALSE`

```{r}
#need to scale because we saw that the winpercent values are on 1-100 while others are 0-1
pca <- prcomp(candy,scale=TRUE)
summary(pca)
```

my PCA plot (a.k.a.) PC1 vs PC2 score plot.

```{r}
plot(pca$x[,1],pca$x[,2],col=my_cols,pch=16)
```
ggplot to make things prettier

```{r}
#make a data frame for ggplotting
my_data<-cbind(candy,pca$x[,1:3])
```
```{r}
p<-ggplot(my_data) + aes(x=PC1,y=PC2,size=winpercent/100,text=rownames(my_data),label=rownames(my_data)) + geom_point(col=my_cols)

p
```
```{r}
#use ggrepel to add labels
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
#interactive plot with plotly
library(plotly)
#ggplotly(p)
#hashtagging plotly out for pdf purposes
```

```{r}
#let's see each candy type contribution to PC1
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, pluribus. negative is chocolate, bar, priceprecent, winpercent, etc. These make sense because we see these based off of where they lie on PC1 in the graph.


